Individual and Differential Harm in Redistricting

### 2022-03-29 paper/paper.Rmd
get compile working

First, we consider the case of Minichusetts.
Modeled on Massachusetts, Minichusetts contains a remarkable lack of partisan segregation.
This high level of integration, combined with Democrats receiving the majority of votes, leads to a Republican inability to win any seats under recent historical voting patterns \[@duch:18\].
For existing measures, this creates a conflict: Republicans win about 40% of the vote, but zero of the nine seats.
The Efficiency Gap would correctly identify that the enacted map favors Democrats, whereas both Partisan Bias and Mean-Median indicate a pro-Republican bias (ref planscore).

@@

Not all cases are as simple as those where one party wins all the seats.
Mininois also has a Democratic majority, but the voters are no longer integrated.
In a familiar pattern of partisan sorting, Democrats are sorted asymmetrically into a single population-dominating city whereas Republicans dominate the less ideologically segregated rural regions \[@chen:rodd:13\] \[@enos:brown:21\]. This allows for significantly greater variation in the expected partisanship of the districts drawn.

@@

A Democratic gerrymander in Mininois effectuates Democratic control by splitting its city into with the blades of a pinwheel reaching out into the more Republican, rural remainder of the state..
If we consider the two plans in Figure TODO, we see that the Democratic gerrymander would likely elect zero Republicans and three Democrats, whereas the fair map would likely elect two Democrats and one Republican.
Consider the precincts \[4, 1\] (maybe highlighted in a color).
\[4,1\] is placed in a Republican district in the fair map, while it is in a Democratic district in the gerrymandered map.
This precinct has 2 Republicans who are not represented by their co-partisan in the gerrymandered plan, but would have been represented by a co-partisan in the fair plan.
Conversely, the 1 Democrat is represented by their co-partisan in the gerrymandered plan, but not the fair plan.
Each of these three voters can show a counterfactual where they would be better represented and thus have a claim to harm.
Yet, if we instead consider any voter who lives in the (other color) \[1,1\] precinct, those voters must always be in a Democratic district given the distribution of voters.
While the Republicans in that precinct are not represented by their co-partisan, they never would be.
Those Republicans can't be individually harmed by redistricting.

@@

Figure \ref{fig:mini-harm} illustrates this measure for a set of hypothetical gerrymanders, comparing to the set of optimal plans from Figure \ref{fig:mini-util-opt}.
As might be intuitively expected, total harm to voters is concentrated in areas whose party representation flips from the gerrymander to the counterfactual plan.

In the districting plan examples of Figure \ref{fig:mini-harm}, the differential harm of the Democratic gerrymander of Mininois is $-0.519$, the differential harm of the Democratic gerrymander of Minissouri is $-0.415$, and the differential harm of the Republican gerrymander of Minissouri is $0.450$, where positive values indicate Republican bias.
In other words, the average Republican voter in Mininois is 51.9% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the utility-maximizing plan in Figure \ref{fig:mini-util-opt}(a).

| State      | Plan                   | Total Harm | Differential Harm | Efficiency Gap | Partisan Bias | Mean-Median | Declination |
|------------|------------------------|------------|-------------------|----------------|---------------|-------------|-------------|
| Mininois   | "Fair"                 |            |                   |                |               |             |             |
| Mininois   | Democratic Gerrymander |            | -0.519            |                |               |             |             |
| Minissouri | Fair                   |            |                   |                |               |             |             |
| Minissouri | Democratic Gerrymander |            | -0.415            |                |               |             |             |
| Minissouri | Republican Gerrymander |            | 0.450             |                |               |             |             |

@@

Fairness is a strict condition; any real-life plan will never be exactly fair.
To quantify deviations from fairness, we introduce the concept of *differential harm*, defined as $$
  \Delta_{G_k-G_l}H(q) \dfeq \E[H(q)\mid I\in G_k] - \E[H(q)\mid I\in G_l],
$$ for two groups $G_k$ and $G_l$.
So $\Delta_{G_k-G_l}H(q)$ will be positive if voters in group $G_k$ are harmed at a higher rate than voters in group $G_l$, indicating that plan $q$ favors $G_l$ over $G_k$.
Using our definition of fairness, $\Delta_{G_k-G_l}H(q)=0$ for all $G_k,G_l\in\Groups$ if and only if $q$ is fair.
<!--
From here onward we will restrict our attention to the case of two groups---concretely, $\Groups=\{\dem,\rep\}$---and simply write $\Delta H(q)=\Delta^\dem_\rep H(q)$.
But the same concepts and definitions naturally carry over to the multigroup setting.
-->

@@

More complicated election outcome models are certainly possible and straightforward to use with the measures proposed here.
For example, one could model district deviations as a systematic national shift plus a local idiosyncratic effect, as is done in the Stochastic Uniform Partisan Swing model of @gelman1994unified.
For ease and efficiency of computation, we stick to the simple proportionate uniform shift model here.

@@

```{=tex}
\begin{restatable}{prop}{propfairness} \label{prop:fairness}
Suppose $\Groups=\{\dem,\rep\}$. If turnout is equal across districts, then for a specific districting plan $q\in\Plans$,
\begin{align*}
    \Delta_{\rep-\dem} H(q) = \E[\seats(\dem, V, \tilde Q)] - \E[\seats(\dem, V, q)].
\end{align*}
\textbf{CM note: define $\seats$ or replace with sum.}

\end{restatable}
```
# Application to Partisan Gerrymandering in New Jersey {#sec:nj}

@@

```{=tex}
\vspace{1in}
\appendix
\renewcommand\thefigure{\thesection.\arabic{figure}}
\setcounter{figure}{0}
```
# Proofs of Propositions {#app:proofs}

\propfairness

```{=tex}
\begin{proof}
\end{proof}
```
# Computing Harm Measures from Precinct-level Election Data {#app:formulas}

>>>>

First, we consider the case of Minichusetts.
Modeled on Massachusetts, Minichusetts contains a remarkable lack of partisan segregation.
This high level of integration, combined with Democrats receiving the majority of votes, leads to a Republican inability to win any seats under recent historical voting patterns [@duch:18].
For existing measures, this creates a conflict: Republicans win about 40% of the vote, but zero of the nine seats.
The Efficiency Gap would correctly identify that the enacted map favors Democrats, whereas both Partisan Bias and Mean-Median indicate a pro-Republican bias (ref planscore).

@@

Not all cases are as simple as those where one party wins all the seats.
Mininois also has a Democratic majority, but the voters are no longer integrated.
In a familiar pattern of partisan sorting, Democrats are sorted asymmetrically into a single population-dominating city whereas Republicans dominate the less ideologically segregated rural regions [@chen:rodd:13; @enos:brown:21].
This allows for significantly greater variation in the expected partisanship of the districts drawn.

@@

A Democratic gerrymander in Mininois effectuates Democratic control by splitting its city into with the blades of a pinwheel reaching out into the more Republican, rural remainder of the state..
If we consider the two plans in Figure TODO, we see that the Democratic gerrymander would likely elect zero Republicans and three Democrats, whereas the fair map would likely elect two Democrats and one Republican.
Consider the precincts $(4, 1)$ (maybe highlighted in a color).
$(4,1)$ is placed in a Republican district in the fair map, while it is in a Democratic district in the gerrymandered map.
This precinct has 2 Republicans who are not represented by their co-partisan in the gerrymandered plan, but would have been represented by a co-partisan in the fair plan.
Conversely, the 1 Democrat is represented by their co-partisan in the gerrymandered plan, but not the fair plan.
Each of these three voters can show a counterfactual where they would be better represented and thus have a claim to harm.
Yet, if we instead consider any voter who lives in the (other color) $(1, 1)$ precinct, those voters must always be in a Democratic district given the distribution of voters.
While the Republicans in that precinct are not represented by their co-partisan, they never would be.
Those Republicans can't be individually harmed by redistricting.

@@

Figure \ref{fig:mini-harm} illustrates this measure for a set of hypothetical gerrymanders, comparing to the set of optimal plans from Figure .
As might be intuitively expected, total harm to voters is concentrated in areas whose party representation flips from the gerrymander to the counterfactual plan.

In the districting plan examples of Figure \ref{fig:mini-harm}, the differential harm of the Democratic gerrymander of Mininois is $-0.519$, the differential harm of the Democratic gerrymander of Minissouri is $-0.415$, and the differential harm of the Republican gerrymander of Minissouri is $0.450$, where positive values indicate Republican bias.
In other words, the average Republican voter in Mininois is 51.9% more likely to be represented by a Democrat than an average Democrat is to be represented by a Republican, compared to the utility-maximizing plan in Figure.

@@

Fairness is a strict condition; any real-life plan will never be exactly fair.
To quantify deviations from fairness, we introduce the concept of *differential harm*, defined as $$
  \Delta_{G_k-G_l}H(q) \dfeq \E[H(q)\mid I\in G_k] - \E[H(q)\mid I\in G_l],
$$ for two groups $G_k$ and $G_l$.
So $\Delta_{G_k-G_l}H(q)$ will be positive if voters in group $G_k$ are harmed at a higher rate than voters in group $G_l$, indicating that plan $q$ favors $G_l$ over $G_k$.
Using our definition of fairness, $\Delta_{G_k-G_l}H(q)=0$ for all $G_k,G_l\in\Groups$ if and only if $q$ is fair.

@@

More complicated election outcome models are certainly possible and straightforward to use with the measures proposed here.
For example, one could model district deviations as a systematic national shift plus a local idiosyncratic effect, as is done in the Stochastic Uniform Partisan Swing model of @gelman1994unified.
For ease and efficiency of computation, we stick to the simple proportionate uniform shift model here.
In Section \ref{sec:boston}, however, we consider a more complicated model for primary elections.

@@

```{=tex}
\begin{proprep} \label{prop:fairness}
Suppose $\Groups=\{\dem,\rep\}$. If turnout is equal across districts, then for a specific districting plan $q\in\Plans$,
\begin{align*}
    \Delta_{\rep-\dem} H(q) = \E[\seats(\dem, V, \tilde Q)] - \E[\seats(\dem, V, q)].
\end{align*}
\textbf{CM note: define $\seats$ or replace with sum.}
\end{proprep}
\begin{proof}
\end{proof}
```
# Application to Partisan Gerrymandering in New Jersey {#sec:nj}

@@

```{=tex}
\vspace{1in}
\appendix
\renewcommand\thefigure{\thesection.\arabic{figure}}
\setcounter{figure}{0}
```

# Computing Harm Measures from Precinct-level Election Data {#app:formulas}
